---
title: "Clean Records"
author: "steppe"
date: "2023-12-04"
output: html_document
---

Here we perform a number of steps which we use to ensure that the records going into the modelling process are as clean as can be. 


Species vary in their responses to disturbances such as fire. 
More importantly, ecosystems differ in their responses to fire based both upon intrinisic characteristics, but especially land management history. 
Many species are unable to recover post wildfire in the West due to the rapid increase in cover of inasive annuals, especially grasses such as Cheatgrass and Medusa-Head. 
Using the NFIC fire perimeters data set, we flag each record which has experienced a wild lands burn, and determine which AIM plots have been located in burn scars.

To do this, we first (1) identify all AIM plots with presences of the species. 
We then (2) combine the plot based and plot-free occurrence data to create occurrence data set. 
3) All remaining AIM plots within 5 miles of the occurrence data set, and within 10% of the range of all environmental variables are identified. 
4) These remaining AIM plots without presences are coded as absences, and used as '0' for the records. 

```{r load libraries}
library(tidyverse)
library(sf)
library(terra)
library(leaflet)
source('functions.R')
```


```{r Import Predictor Stack for outlier detection, eval = F}

p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa/bio10'
files <- file.path(p, list.files(p, recursive = TRUE))

# bio10 <- vrt(files)
# bio1 <- vrt(files)
# bio12 <- vrt(files)
# bio4 <- vrt(files)
# bio19 <- vrt(files)
names(bio1) <- 'bio1'
names(bio10) <- 'bio10'
names(bio12) <- 'bio12'
names(bio4) <- 'bio4'
names(bio19) <- 'bio19'

bio <- c(bio1, bio10, bio12, bio4, bio19)
# filter(thinned) |> 
#   pull(taxon) |> 
#   unique() |>
#  write.table('../data/raw/species.txt', row.names = FALSE)

rm(bio1, bio10, bio12, bio4, bio19, p, files)
```

```{r manually review outliers,  eval = F}

st <- tigris::states() %>% 
    filter(DIVISION %in% c(8, 9), ! STUSPS %in% c('AK', 'HI')) %>% 
  st_transform(4326) 

st_crs(thinned)

fk <- st_intersection(filter(st, STUSPS %in% c('OR', 'ID', 'WY', 'CO', 'WA')) %>% 
                  st_transform(5070),  bio_v1) %>% 
  pull(ID)

paste0(fk, collapse = "", sep = ', ')

# merge all stephanomeria subspecies with species
# Dudleya cymosa ssp. cymosa merge with parent

pts1 <- filter(thinned, taxon == 'Penstemon palmeri')
bio_v <- extract(bio, pts1)
bio_v1 <- bind_cols(bio_v, pts1) %>% 
  st_as_sf() 

OUTPUT <- quantile_flagger(bio_v1, 0.015, flag_threshold = 0.2) %>% 
  st_transform(4326) %>% 
  mutate(LONG_WGS = st_coordinates(.)[,1], 
         LAT_WGS = st_coordinates(.)[,2]) 

OUTPUT <- filter(OUTPUT, ! ID %in% c(3:8, 26, 34:35, 47:55, 73:77, 92:94, 98:99, 102:103, 115, 122, 129, 139, 143, 149, 168, 195, 199, 209, 214, 216, 232, 236, 247, 256, 264, 267, 268, 273, 281, 284, 285, 297, 302:303, 305, 312, 318, 326:327, 333, 343, 348, 349:350, 360, 373, 390, 399, 404, 406, 419, 423, 437:438, 447, 454, 458, 462, 470, 472, 474, 494, 509, 511, 514, 536, 539, 544, 558, 570, 572, 573, 576, 610, 611, 616, 623, 629, 632:633, 635, 651, 663, 684, 696, 711, 732, 733, 757, 767, 812, 814, 816, 819, 821, 828, 836, 849, 850, 856, 858, 874, 890, 901, 910, 939, 946, 958, 962, 966:967, 971, 981, 989, 997, 1007, 1010, 1016, 1020, 1031, 1054))

ggplot() +  
  geom_sf(data = OUTPUT, aes(color = OutlierFlag)) +
  geom_sf_text(data = OUTPUT, aes(label = ID)) +
  geom_sf(data = st, fill = NA) 

leaflet(data = filter(OUTPUT, ID < 1100)) %>%  addTiles() %>%
   addMarkers(~LONG_WGS, ~LAT_WGS,  popup = ~as.character(ID), label = ~as.character(ID))

rm(bio_v, bio_v1, OUTPUT, pts1)
```


Every single taxon was individually mapped, and had their records investigated. Points which were in the 1.5% most extreme values for at least two values of BIO1, BIO4, BIO10, BIO12, BIO19, and for distance from the nearest population were color coded for investigations. 

```{r subset occurrence records for final clean batch}
thinned <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE)
thinned2 <- st_read('../data/raw/occurrence/thinned/thinned_occurrences2.shp', quiet = TRUE)
pts2remove <- read.delim('../data/raw/species_manual_cleaning.txt', header = FALSE)
colnames(pts2remove) <- 'Species'

pos <- str_extract_all(pts2remove$Species, "\\([^()]+\\)")
pos <- lapply(pos, function(x){substring(x, 2, nchar(x)-1)})
exact <- lapply(pos, function(x){suppressWarnings({as.numeric(sub(' ' , '', unlist(strsplit(x,","))))})})
ranges <- lapply(pos, colon_balloon)

to_remove <- Map(c, exact, ranges)  
to_remove <- lapply(to_remove, na.omit)
to_remove <- lapply(to_remove, sort)

rm(exact, ranges)

thinned <- bind_rows(thinned, thinned2) %>% 
  mutate(taxon = str_replace(taxon, 'Acmispon humistratus', 'Acmispon brachycarpus'), 
         taxon = str_replace(taxon, 'Adenostoma fasiculatum', 'Adenostoma fasciculatum'),
         taxon = str_replace(taxon, 'Brodiaeae elegans', 'Brodiaea elegans')
         ) %>% 
  filter(! taxon %in% c('Ceanothus pauciflorus', 'Cirsium acuminata', 'Mirabilis lanata')) %>% 
  arrange(taxon)

thinned <- thinned %>% 
  group_by(taxon) %>% 
  mutate(ID = 1:n(), .before = taxon)

splicies <- split(thinned, f = thinned$taxon)

splicies_l <- vector(mode = 'list', length = length(splicies))
for (i in seq_along(splicies)){
  if(to_remove[[i]] == 0){
    splicies_l[[i]] <- splicies[[i]]
  } else{
    splicies_l[[i]] <- splicies[[i]][-to_remove[[i]], ]
  }
}

splicies <- bind_rows(splicies_l)

rm(pos, pts2remove, splicies_l, thinned, thinned2, i, colon_balloon, to_remove)
# now combine synonyms
```

```{r load nifc}

pNIFC <- '/media/steppe/hdd/Geospatial_data/NIFC_perimeters'
NIFC <- st_read(file.path(pNIFC, 'InterAgencyFirePerimeterHistory_All_Years_View.shp')) |>
  select(OBJECTID, FIRE_YEAR, INCIDENT) |>
  st_make_valid()

domain <- rast(nrows = 1, ncols = 1) # create a big empty raster, you can go in through sf too. 
ext(domain) <- c( -125.5, -100, 27,  50) # set the extent
domain <- as.polygons(domain) |>  # convert to vector data
  st_as_sf() |> # convert to simple feature
  st_transform(4326) %>% 
  st_bbox()

NIFC <- st_crop(NIFC, domain)
print(object.size(NIFC), units = "GiB", standard = "IEC") # 600971416
NIFC <- st_simplify(NIFC, dTolerance = 90)
print(object.size(NIFC), units = "GiB", standard = "IEC")

ggplot() + 
  geom_sf(data = filter(NIFC, INCIDENT == 'DEL VENTURI' & FIRE_YEAR == 2003))

102910208 / 600971416
```

```{r extract most recent fire}

```


We will identify environmental parameters associated with species occurrences, if absences and 'pseudo absences' are very similar to these values, we will discard certain records within this space. 

```{r identify similar plots}

p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa'

paths <- file.path(p, list.files(p, recursive = F))
files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
names(files) <- basename(paths)

bio <- lapply(files, vrt)
bio <- rast(bio)

abvi <- filter(splicies, taxon == 'Eriophyllum lanatum') |>
  st_transform(5070)

ggplot(abvi) + 
  geom_sf(aes(color = ID)) + 
  geom_sf(data = filter(abvi, ID == 909), color = 'red')

test <- extract(bio, abvi)
test <- test %>% 
  select(-gdgfgd10, -gdgfgd5, -swe, -scd, -ID) %>% 
  drop_na()

test1 <- apply(test, MARGIN = 2, FUN = scales::rescale, to = c(0,1))

res.pca <- prcomp(test1)
factoextra::fviz_eig(res.pca)

factoextra::fviz_pca_ind(res.pca,
             col.ind = "cos2", # Color by the quality of representation
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )

factoextra::fviz_pca_var(res.pca,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )


```

```{r prepare data for modelling}

```

```{r model}

```

```{r evaluate models}

```

```{r export model summaries}

```

```{r subset occurrence records to reflect fire history}

```

