---
title: "Clean Records"
author: "steppe"
date: "2023-12-04"
output: html_document
---

Here we perform a number of steps which we use to ensure that the records going into the modelling process are as clean as can be. 


Species vary in their responses to disturbances such as fire. 
More importantly, ecosystems differ in their responses to fire based both upon intrinisic characteristics, but especially land management history. 
Many species are unable to recover post wildfire in the West due to the rapid increase in cover of inasive annuals, especially grasses such as Cheatgrass and Medusa-Head. 
Using the NFIC fire perimeters data set, we flag each record which has experienced a wild lands burn, and determine which AIM plots have been located in burn scars.

To do this, we first (1) identify all AIM plots with presences of the species. 
We then (2) combine the plot based and plot-free occurrence data to create occurrence data set. 
3) All remaining AIM plots within 5 miles of the occurrence data set, and within 10% of the range of all environmental variables are identified. 
4) These remaining AIM plots without presences are coded as absences, and used as '0' for the records. 

```{r load occurrence records}

```

```{r load nifc}

```

```{r extract most recent fire}

```

```{r identify similar plots}

```

```{r prepare data for modelling}

```

```{r model}

```